---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

#import file containing variable dictionaries to  make into markdown tables
library(readxl)
library(knitr)
file_path <- "inst/exdata/kidneyoutcomes_definitions.xlsx"
```

# kidneyoutcomes: Kidney Outcomes Analysis for Transplantation

<!-- badges: start -->
<!-- badges: end -->

The goal of **kidneyoutcomes** is to processes serum creatinine data to identify Acute Kidney Injury (AKI) events and progression to Chronic Kidney Disease (CKD) stages within a defined baseline context. This package was developed for use in islet transplantation but may be adaptable to other transplant settings where measurement of serum creatinine is frequent. Read on for a summary of the implementation of the logic that was used to define AKI and CKD in this setting. 

## Installation

You can install the development version of kidneyoutcomes from [GitHub](https://github.com/) with:

``` r
# Install devtools if not already installed
install.packages("devtools")

# Install kidneyoutcomes
devtools::install_github("yourusername/kidneyoutcomes")
```

## Example

This is a basic example of the package useage :

``` r 
library(kidneyoutcomes)
library(rio)

#load in your appropriately formatted file containing: patient id, date of transplant (or baseline),date of laboratory result, age at laboratory result, sex and creatinine (mg/dl)  
inputdataframe<-rio::import("path/to/input.csv")

# Run the function
kidneyoutcomes::kidneyoutcomesintx(
  inputdataframe = inputdataframe,
  outputdirectory = "path/to/output",
  datecollocation = 3
)
```
### How to format your input file

Please format your input file that you will load into R to run the function over as below. The definitions of each of these variables are also found below. **Please retain names of columns, pay attention to units and coding for the sex variable**. 

```{r, echo = FALSE, results = 'asis'}
library(readxl)
library(knitr)

# Read the tables
table1 <- read_excel(file_path, sheet = "inputstructure")
table2 <- read_excel(file_path, sheet = "dictionary_input")

# Convert each table to HTML with styling
table1_html <- kable(table1, format = "html", table.attr = 'style="width:100%; font-size:12px; border-collapse:collapse;"')
table2_html <- kable(table2, format = "html", table.attr = 'style="width:100%; font-size:12px; border-collapse:collapse;"')

# Combine the tables side by side with labels
cat('<table style="width:100%; table-layout:fixed; font-size:12px;"><tr>')

# Add labels for each table
cat('<th style="width:50%; text-align:center; font-size:14px;">Input File Structure</th>')
cat('<th style="width:50%; text-align:center; font-size:14px;">Definitions of Variables in Input File</th>')

cat('</tr><tr>')

# Add the tables side by side
cat('<td style="vertical-align: top; width:50%; padding: 10px; font-size:12px;">', table1_html, '</td>')
cat('<td style="vertical-align: top; width:50%; padding: 10px; font-size:12px;">', table2_html, '</td>')

cat('</tr></table>')
```

## Identification of AKI and CKD in the Islet transplanation context

This function calculates the [CKD-EPI 2021](https://www.kidney.org/ckd-epi-creatinine-equation-2021-0) derived eGFR per individual from the serum creatinine, age at lab, and sex.

### 1.Identification of CKD Episodes
- Episodes of CKD are first identified based on the [KDIGO criteria](https://www.sciencedirect.com/science/article/pii/S0085253823007664?via%3Dihub) as sustained eGFR below threshold for at least **90 days**.

### 2. Definition of AKI
- AKI is defined based on the [KDIGO criteria](https://www.sciencedirect.com/science/article/pii/S0085253823007664?via%3Dihub) and implemented using the logic of the algorithm applied in the [UK National Health Service (NHS)](https://www.england.nhs.uk/wp-content/uploads/2014/06/psa-aki-alg.pdf). The reference value depended on the number of days since a previous result:  

  - Creatinine result ≥1.5 times higher than the **median** of all previous creatinine measures in the previous 8–365 days.  
  - Creatinine result ≥1.5 times higher than the **lowest creatinine** from the previous 7 days.  
  - Creatinine >26 mmol/L higher than the **lowest creatinine** from the previous 48 hours.  

- Creatinine results in the first 7 days post-transplant were assessed against the previously derived **baseline creatinine** (12-month pre-transplant average) as the reference value.

### 3. AKI Event Grouping
- If an AKI event was within **7 days** of another AKI event, this was identified as a single (i.e., the same) event.

### 4. AKI Overlapping CKD
- AKI events identified in the previous step were not counted as an AKI event if they overlapped with a date that an identified the start of an episode of CKD.  
- However, AKI events could occur during an episode of CKD.

### 5. Identification of Acute Kidney Disease (AKD)
- **Acute Kidney Disease (AKD)** events were identified as AKI events lasting **>7 consecutive days**.  
- However, the "AKI 1-year count" summary includes any AKD events in that total.
 
### 6. Defining "progression to CKD"
- For this context, the identified progression to sustained Stage 3/3b/4/5 CKD is defined based on the [KDIGO criteria (2024)](https://www.sciencedirect.com/science/article/pii/S0085253823007664?via%3Dihub) with an additional criterion: having a sustained eGFR below threshold for at least 90 days, **and an average eGFR in the last 6 months of follow-up below such thresholds**. 

- The thresholds as defined by [KDIGO](https://www.sciencedirect.com/science/article/pii/S0085253823007664?via%3Dihub) are:
  - **Stage 3**: <60 ml/min/1.73m<sup>2</sup>
  - **Stage 3b**: <45 ml/min/1.73m<sup>2</sup>
  - **Stage 4**: <30 ml/min/1.73m<sup>2</sup>
  - **Stage 5**: <15 ml/min/1.73m<sup>2</sup>

- The additional criterion of an average eGFR in the last 6 months of follow-up below such thresholds is to ensure our definition captured likely **irreversible CKD** stages within the islet transplant setting, in which kidney function can be very dynamic due to immunosuppression.

- In identified progressors, **time to progression** of each CKD stage was noted as the **beginning of the first episode where eGFR remained below threshold for ≥90 days**. 

## Function outputs and suggested further uses
1. Individual dataframes containing:

- **akiyear1**: Summary number of AKI events occurring in year 1 after transplant (or some defined baseline) for each patient ID
- **akiall**: Summary of every AKI event that occurred for the individual, containing the start and end time in days from baseline and the length of the event. If given a value of 0 there were no AKI events for that individual
- **dateprogression_fin**: Summary of AKI events over all follow up, the number of which were >7 days and the identification of CKD progression and timing of progression in days with respect to baseline. (eGFR thresholds of 15, 30, 45, and 60)

Definitions of each of the variables for the data frames are below:


```{r, echo = FALSE, results = 'asis'}
library(readxl)
library(knitr)

# Read the tables
table3 <- read_excel(file_path, sheet = "dictionary_output_aki1year")
table4 <- read_excel(file_path, sheet = "dictionary_output_akiall")
table5 <- read_excel(file_path, sheet = "dictionary_output_progression")

# Convert each table to HTML with consistent column widths
table3_html <- kable(table3, format = "html", table.attr = 'style="width:100%; font-size:12px; border-collapse:collapse;"')
table4_html <- kable(table4, format = "html", table.attr = 'style="width:100%; font-size:12px; border-collapse:collapse;"')
table5_html <- kable(table5, format = "html", table.attr = 'style="width:100%; font-size:12px; border-collapse:collapse;"')

# Combine the tables side by side with labels and smaller text
cat('<table style="width:100%; table-layout:fixed; font-size:12px;"><tr>')

# Add labels for each table with smaller text
cat('<th style="width:33%; text-align:center; font-size:12px;">Definitions of variables in akiyear1 dataframe</th>')
cat('<th style="width:33%; text-align:center; font-size:12px;">Definitions of variables in akiall dataframe</th>')
cat('<th style="width:33%; text-align:center; font-size:12px;">Definitions of variables in progression dataframe</th>')

cat('</tr><tr>')

# Add the tables side by side
cat('<td style="vertical-align: top; width:33%; padding: 10px; font-size:12px;">', table3_html, '</td>')
cat('<td style="vertical-align: top; width:33%; padding: 10px; font-size:12px;">', table4_html, '</td>')
cat('<td style="vertical-align: top; width:33%; padding: 10px; font-size:12px;">', table5_html, '</td>')

cat('</tr></table>')

```
  
2. Summary table in a word document for serum creatinine measurement rates and frequencies called **SystemDate_number_of_creatinine_results_assessed.docx** that will be saved in the output directory you spcified in the function.

The dataframes can be downloaded as separate csv/xlsx files or taken on for further analysis within your R script.  

``` r 
#Download the dateprogression_fin dataframe as a csv file
library(rio)

rio::export(dateprogression_fin,"your/output/directory/aki_and_ckd_progression.csv")

```

## Citation

If you use this package in your research, please cite it as:

> Alice Carr (2025). *kidneyoutcomes: Kidney Outcomes Analysis for Transplantation*.  
> GitHub: <https://github.com/alicelouisejane/kidneyoutcomes>.

You can also retrieve the citation in R using:

```r 
citation("kidneyoutcomes")
```

